\section{Preparation}

\noindent
This chapter covers the following ideas. When you create your lesson plan, it should contain examples which illustrate these key ideas. Before you take the quiz on this unit, meet with another student out of class and teach each other from the examples on your lesson plan. 

\input{03-Patterns/Patterns-objectives}

Here are the preparation problems for this unit.  Problems that come from Schaum's Outlines ``Beginning Linear Algebra'' are preceded by a chapter number. From here on out, many of the problems we will be working on come from Schaum's Outlines.  Realize that sometimes the method of solving the problem in Schaum's Outlines will differ from how we solve the problem in class. The difference is that in Schaum's Outlines they almost always place vectors in rows prior to row reduction, whereas we will be placing vectors in columns. There are pros and cons to both methods. 


\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{c}{Preparation Problems
  (\href{http://ilearn.byui.edu/bbcswebdav/institution/Physical\_Sci\_Eng/Mathematics/Personal\%20Folders/WoodruffB/341/3-Patterns-Preparation-Solutions.pdf}{click
    for handwritten solutions})}
\\
\hline\hline
Day 1&
1f,
2a,
2f,
2i
\\ \hline
Day 2&
2j,
4b,
Chp 4:6
Chp 4:13
\\ \hline
Day 3&
Chp 4:22
Chp 5:18
Chp 6:6
Chp 6:9
\\ \hline
Day 4&
Lesson Plan,
Quiz, Start Project 
\\ \hline
\end{tabular}
\end{center}


\subsection{Webcasts}

\href{http://www.youtube.com/user/bmwoodruff\#g/c/54D7F4A63AE6AE18}{Click
  here for webcasts on YouTube}\footnote{\url{http://www.youtube.com/user/bmwoodruff\#g/c/54D7F4A63AE6AE18}}.  \href{http://ilearn.byui.edu/bbcswebdav/institution/Physical\_Sci\_Eng/Mathematics/Personal\%20Folders/WoodruffB/341/3-Patterns-videos.pdf}{Click
  here for a pdf copy of the papers in the webcasts}.

\subsection{Suggested Problems}

The first three sets of problems are in this book.  The remaining problems come from Schaum's Outlines ``Beginning Linear Algebra'' by Seymour Lipschutz.  
\begin{center}
\begin{tabular}{|l|l|l|l|l|}
\hline
Concept&Suggestions&Relevant Problems\\ \hline
Illustrating Definitions
	&\ref{definition problems}cdf
  &\ref{definition problems}
  \\ \hline
Illustrating Theorems
	& \ref{theorem problems}acdfijl (all are good)  
	&\ref{theorem problems}
	\\ \hline
The huge equivalence
	&\ref{equivalence problems}abe 
	&\ref{equivalence problems}
	\\ \hline
Vector Spaces
	& In Schaum's
	&\ref{vector space problems}
	\\ \hline
Subspaces
	& In Schaum's
	&\ref{subspace problems}
	\\ \hline
Spans
	&In Schaum's
	&\ref{span problems}
	\\ \hline
Basis and Dimension
	&In Schaum's
	&\ref{basis problems}
	\\ \hline
Coordinates
	&In Schaum's
	&\ref{coordinates problems}
	\\ \hline
Linear Independence
	&In Schaum's
	&\ref{independence problems}
	\\ \hline
\end{tabular}
\end{center}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../linear-algebra"
%%% End: 
